---
title: "March Madness 2017"
output: html_notebook
---

This notebook uses historical March Madness data to predict the results of the 2017 tournament.

## Reading the data
The data comes in the following files:
```{bash}
ls | grep csv$
```
When training the model, we mostly care about using the detailed results to develop features. The detailed data, both tournament and regular season, contains game by game information as follows:
```{r Setup}
# Setup
library(data.table)
library(tidyverse)
```

```{r Read Data}
regularSeason <- fread("RegularSeasonDetailedResults.csv")
tournament <- fread("TourneyDetailedResults.csv")
head(regularSeason)
```
## Tidying the data
To generate training and testing data, we need to work with data tables with the same formatting as the sample submission, which is formatted as follows:
```{r}
sampleSub <- fread("sample_submission.csv")
head(sampleSub)
```
We need to get ids for all the games that contain the season, followed by the team with the lower team id, then the team with the higher team id. 
The label is 1 when the team with the lower team id wins, and 0 otherwise. To create this data table, we use the following function:
```{r}
# Converts detailed data to submission data
convertToSub <- function(dt) {
    dt <- copy(dt)
    dt[, id := paste(Season, ifelse(Wteam > Lteam, Lteam, Wteam), ifelse(Wteam < Lteam, Lteam, Wteam), sep = "_")]
    dt[, label := as.numeric(Wteam < Lteam)]
    dt[, .(id, label)]
}

labeledSeason <- convertToSub(regularSeason)
labeledTournament <- convertToSub(tournament)
head(labeledSeason)
head(labeledTournament)
```
Now that we have labeled data that we can add features to, we use the historical data to generate features.

## Working with the Historical Data















